TiChri: Matrix correction integration in patRoon

### 2024-01-17 dev/TiChri_integration_patRoon.Rmd
Added entries to gitignore

dbis <- all_db[grepl("IS", all_db$tag), ]
cols <- c("name", "formula", "mass", "rt")
dbis <- dbis[, cols, with = FALSE]
```

@@

```{r tics, message=FALSE, fig.height=9, fig.cap="Total ion chromatogram (TIC) of all analyses by replicate group name."}
ms <- MassSpecData$new(selected_files)
ms$add_replicate_names(anaInfo$group)
ms$add_blank_names(anaInfo$blank)
ms$plot_tic(levels = 1, colorBy = "replicates")
```

@@

```{r data-process, message=FALSE, results='markup'}
ms$find_features(
  Settings_find_features_openms(
    noiseThrInt = 1000,
    chromSNR = 3,
    chromFWHM = 7,
    mzPPM = 15,
    reEstimateMTSD = TRUE,
    traceTermCriterion = "sample_rate",
    traceTermOutliers = 5,
    minSampleRate = 1,
    minTraceLength = 4,
    maxTraceLength = 70,
    widthFiltering = "fixed",
    minFWHM = 4,
    maxFWHM = 35,
    traceSNRFiltering = TRUE,
    localRTRange = 0,
    localMZRange = 0,
    isotopeFilteringModel = "none",
    MZScoring13C = FALSE,
    useSmoothedInts = FALSE,
    extraOpts = NULL,
    intSearchRTWindow = 3,
    useFFMIntensities = FALSE,
    verbose = FALSE
  )
)

ms$group_features(
  Settings_group_features_openms(
    rtalign = FALSE,
    QT = FALSE,
    maxAlignRT = 8,
    maxAlignMZ = 0.01,
    maxGroupRT = 8,
    maxGroupMZ = 0.01,
    extraOptsRT = NULL,
    extraOptsGroup = NULL,
    verbose = FALSE
  )
)

@@

@@

```{r internal-standards-eic, fig.height=9, fig.cap="Extracted ion chromatogram (EIC) of four spiked internal standards per analysis replicate group.", message=FALSE}
ms$plot_eic(mass = dbis[c(3, 4, 7, 8), c("name", "mass"), with = FALSE], colorBy = "replicates+targets", legendNames = TRUE, ppm = 20)
```

@@

      mp <- vapply(seq_len(nrow(ticX)), function(j, z) {
        rt <- ticX$rt[j]
        rtr <- c(rt - rtWindow, rt + rtWindow)
        intB <- mean(z$int[.trimVector(z$rt, rtr[1], rtr[2])])
        # eq. 6 from 10.1021/acs.analchem.1c00357
        (ticX$int[j] / intB) * -1
      }, z = z, 0)

@@

  out <- data.table(
    "feature" = dt$feature,
    "group" = dt$group,
    "analysis" = dt$analysis,
    "rt" = dt$rt,
    "intensity" = dt$intensity,
    "scale" = NA_real_,
    "mp" = NA_real_,
    "corrected" = NA_real_
  )

  for (i in seq_len(nrow(out))) {
    mp <- ticMps[[out$analysis[i]]]

@@

    if (!any(selBlkInts)) {
      warning("Feature", out$feature[i], " from analysis ", out$analysis[i], " does not have blank representation!")
      next
    }

@@

    #based on eq. 8 from 10.1021/acs.analchem.1c00357
    out$corrected[i] <- out$intensity[i] * ((1 - out$mp[i] + 1) - 1)
  }

@@

corrInts <- calculateCorrectedIntensities(dt, anaInfo, ticMps, rtWindow = 4)

@@

>>>>

dbis <- all_db[grepl("IS", all_db$tag), ]
cols <- c("name", "formula", "mass", "rt")
dbis <- dbis[, cols, with = FALSE]
dbis <- dbis[c(3, 4, 7, 8), ]
```

@@

```{r tics, message=FALSE, fig.cap="Total ion chromatogram (TIC) of all analyses by replicate group name."}
ms <- MassSpecData$new(selected_files)
ms$add_replicate_names(anaInfo$group)
ms$add_blank_names(anaInfo$blank)
ms$plot_tic(levels = 1, colorBy = "replicates")
```

@@

```{r data-process, message=FALSE, results='markup'}
ms$find_features(
  Settings_find_features_openms(
    noiseThrInt = 1000,
    chromSNR = 5,
    chromFWHM = 5,
    mzPPM = 15,
    reEstimateMTSD = TRUE,
    traceTermCriterion = "sample_rate",
    traceTermOutliers = 5,
    minSampleRate = 1,
    minTraceLength = 4,
    maxTraceLength = 70,
    widthFiltering = "fixed",
    minFWHM = 4,
    maxFWHM = 35,
    traceSNRFiltering = TRUE,
    localRTRange = 0,
    localMZRange = 0,
    isotopeFilteringModel = "none",
    MZScoring13C = FALSE,
    useSmoothedInts = FALSE,
    extraOpts = NULL,
    intSearchRTWindow = 3,
    useFFMIntensities = FALSE,
    verbose = FALSE
  )
)

ms$group_features(
  Settings_group_features_openms(
    rtalign = TRUE,
    QT = FALSE,
    maxAlignRT = 10,
    maxAlignMZ = 0.02,
    maxGroupRT = 8,
    maxGroupMZ = 0.01,
    extraOptsRT = NULL,
    extraOptsGroup = NULL,
    verbose = FALSE
  )
)

@@

```{r internal-standards-eic, fig.height=7, fig.cap="Spiked internal standards after feature finding and grouping.", message=FALSE}
ms$plot_groups(ms$get_internal_standards()[,c("group", "name"), with = FALSE], legendNames = TRUE, colorBy = "targets+replicates")
```

@@

```{r internal-standards-eic-old, include=FALSE, fig.height=7, fig.cap="Extracted ion chromatogram (EIC) of four spiked internal standards per analysis replicate group.", message=FALSE}
# ms$get_internal_standards(average = FALSE)
# ms$plot_eic(mass = dbis[, c("name", "mass"), with = FALSE], colorBy = "replicates+targets", legendNames = TRUE, ppm = 20)
# ms$plot_eic(mass = dbis[c(3, 4, 7, 8), c("name", "mass"), with = FALSE], colorBy = "replicates+targets", legendNames = TRUE, ppm = 20)
```

@@

      mp <- vapply(seq_len(nrow(ticX)), function(j, z) {

        rt <- ticX$rt[j]

        rtr <- c(rt - rtWindow, rt + rtWindow)

        intB <- mean(z$int[.trimVector(z$rt, rtr[1], rtr[2])])

        intX <- mean(ticX$int[.trimVector(ticX$rt, rtr[1], rtr[2])])

        # eq. 6 from 10.1021/acs.analchem.1c00357
        (intX / intB) * -1

        # alternative using the intensity at the exact rt
        # (ticX$int[j] / intB) * -1

      }, z = z, 0)

@@

  out <- data.table(
    "feature" = dt$feature,
    "group" = dt$group,
    "analysis" = dt$analysis,
    "rt" = dt$rt,
    "intensity" = dt$intensity,
    "ticmp" = NA_real_,
    "scale" = NA_real_,
    "mp" = NA_real_,
    "corrected" = NA_real_
  )

  for (i in seq_len(nrow(out))) {

    mp <- ticMps[[out$analysis[i]]]

@@

    if (!any(selBlkInts)) {

      # What to do?
      # Find another feature in the near retention time?
      # Pre-calculate retention time and mass based scales?

      warning("Feature", out$feature[i], " from analysis ", out$analysis[i], " does not have blank representation!")

      next
    }

    out$ticmp[i] <- mean(mp$mp)

@@

    #based on eq. 8 from 10.1021/acs.analchem.1c00357
    out$corrected[i] <- out$intensity[i] * ((1 - out$mp[i] + 1) - 1)

  }

@@

rpls <- ms$get_replicate_names()

corrInts <- calculateCorrectedIntensities(dt, anaInfo, ticMps, rtWindow = 4)
corrInts$replicate <- rpls[corrInts$analysis]

@@

```{r, include=FALSE}

dt2 <- ms$get_features(mass = db, sec = 15, ppm = 10)
ms$plot_groups_overview(groups = dt2, legendNames = TRUE)

ms$plot_eic(mass = db, legendNames = TRUE)

corrInts2 <- calculateCorrectedIntensities(dt2, anaInfo, ticMps, rtWindow = 4)

library(ggplot2)

ggplot(corrInts, aes(x=rt, y=scale, shape=replicate, color=group)) +
  geom_point()

ggplot(corrInts, aes(x=rt, y=ticmp, shape=replicate, color=group)) +
  geom_point()

ggplot(corrInts, aes(x=mp, y=ticmp, shape=replicate, color=group)) +
  geom_point()

fig <- plot_ly(corrInts, x = corrInts$rt)

fig <- fig %>% add_trace(corrInts,
  x = corrInts$rt,
  y = corrInts$scale,
  color = corrInts$replicate,
  name = corrInts$replicate,
  legendgroup = corrInts$replicate,
  type = "scatter", mode = "markers",
  marker = list(size = 10)
)

fig2 <- plot_ly(corrInts, x = corrInts$rt)

fig2 <- fig2 %>% add_trace(corrInts,
  x = corrInts$rt,
  y = corrInts$ticmp,
  color = corrInts$replicate,
  name = corrInts$replicate,
  legendgroup = corrInts$replicate,
  type = "scatter", mode = "markers",
  marker = list(size = 10),
  showlegend = FALSE
)

subplot(
  list(fig, fig2),
  nrows = 2,
  titleY = TRUE, titleX = TRUE,
  margin = 0.01,
  shareX = TRUE,
  which_layout = "merge"
)

```
